Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-04-14 article_springer/article_springer.tex
Last small changes

Often a magnitude, or mark can be assigned to each event in the renewal
process, such as for earthquakes, solar flares or neuron voltages. The
Peaks-Over-Threshold model (POT, see e.g.~Coles, 2001) applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution
to the threshold exceedances. A commonly made assumption in POT models
is that times between events are either fixed or light-tailed, and this
entails that the threshold crossing times form a Poisson process (``On
the Exceedance Point Process for a Stationary Sequence,'' 1988). Then as
one increases the threshold and thus decreases the threshold crossing
probability \(p_{\ell}\), the Poisson process is rarefied, i.e.~its
intensity decreases \emph{linearly} with \(p_{\ell}\) (see e.g.~Beirlant
et al., 2006).

As will be shown below, in the heavy-tailed waiting time scenario
threshold crossing times form a \emph{fractional Poisson process}
(Laskin, 2003; Meerschaert et al., 2011), which is a renewal process
with Mittag-Leffler distributed waiting times. The family of
Mittag-Leffler distributions nests the exponential distribution (Haubold
et al., 2011), and hence the fractional Poisson process generalizes the
standard Poisson process. Again as the threshold size increases and the
threshold crossing probability \(p_{\ell}\) decreases, the fractional
Poisson process is rarefied: The scale parameter of the Mittag-Leffler
inter-arrival times of threshold crossing times increases, but
\emph{superlinearly}; see the Theorem below.

@@

In the following we denote with \(x_L \in [-\infty, +\infty)\) and
\(x_R \in (-\infty, +\infty]\) the left and right endpoint of the
distribution of \(J\). We assume that the \(k\)-th magnitude \(J_k\)
occurs at time \(T_k = W_1 + \ldots + W_k\). Based on an MRP, we define
the Continuous Time Random Exceedance model (CTRE) as follows:

@@

\begin{description}
\item[\textbf{Theorem:}]
Let the waiting times \(W_k\) be in the domain of attraction of a
positively skewed sum-stable law with stability parameter
\(0 < \beta < 1\); more precisely, \begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align} for a function \(b(n)\) which is regularly varying at
\(\infty\) with parameter \(1/\beta\), and where
\(\mathbf E[\exp(-sD)] = \exp(-s^\beta)\). Write
\(p_{\ell} := \mathbf P(J > \ell)\). Then the weak convergence \[
\frac{T(\ell)} {b(1/p_{\ell})} \to Z_\beta \quad \text{ as } \quad \ell \uparrow x_R
\] holds, where the Mittag-Leffler random variable \(Z_\beta\) is
defined on the positive real numbers via \[
\mathbf E[\exp(-sZ_\beta)] = \frac{1}{1+s^\beta}.
\]
\end{description}

\emph{Proof of Theorem:} Due to assumption \eqref{assumptionJs}, it
follows \begin{align*}
    P\left(\frac{M(c)-d(c)}{a(c)} \leq x\right) \rightarrow F(x) \text{ as } c \rightarrow \infty,
    \end{align*} where \(x\) is any value from the support of \(F\).
Then it follows, \begin{align*}
      F(xa(c)+d(c))^{\lfloor c \rfloor} &\rightarrow F(x) \text{ as } c \rightarrow \infty,\\
      \Rightarrow  \lfloor c \rfloor \log F(\ell_c)  &\rightarrow \log F(x)\text{ as } c \rightarrow \infty,
    \end{align*} with \(\ell_c:=xa(c)+d(c)\). Furthermore, since
\(\log(1-x) \sim -x\) for small \(x\) it follows \begin{align}
      c \cdot p(\ell_c) \rightarrow -\log F_A(x)  \text{ as } c \rightarrow \infty,\label{eq:lplconv}
    \end{align} with \(p(\ell_c):= \overline{F}(\ell_c)\). Since
\(\tau(\ell_c) \sim Geo(p(\ell_c))\) and \eqref{eq:lplconv} it follows
that \(\tau(\ell_c)/c\) converges to an Exponential distributed random
variable, \begin{align*}
    \frac{\tau(\ell_c)}{c} \Rightarrow E  \text{ as } c \rightarrow \infty
    \end{align*} with parameter \(\lambda:=-\log F_A(x)\). Due to
\begin{align*}
    S(c):= \sum_{i=1}^{\lfloor c \rfloor} \frac{W_i}{b(c)} \Rightarrow D \ \text{ as } c \rightarrow \infty,
    \end{align*} it follows with Gnedenkos transfer theorem (see
Gnedenko (1983)), that \begin{align}
    \sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(\ell_c)}  \Rightarrow Z\ \text{ as }  c \rightarrow \infty, \label{TheoGutArg}
    \end{align} where the distribution of \(Z\) has the characteristic
function \begin{align}
    \Psi(t) &= \int_0^{\infty} (\Phi_D(t))^{y} F_E(dy) \nonumber\\
    &= \int_0^{\infty} \frac{1}{1-\log(\Phi_D(t))/\lambda} \nonumber\\
    &= \frac{1}{1+s^{\beta}\lambda^{-1}}
    \end{align} where \(\Phi_D\) resp. \(F_E\) denotes Laplace transform
of \(D\) resp. the distribution function of \(E\). Hence, \(Z\) is
Mittag-Leffler distributed with parameter \(\lambda^{-1/\beta}\). Since
\begin{align*}
    \sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(1/p(\ell_c))} = \left(\sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(c) }\right) \frac{b(c)}{b(1/p(\ell_c))}  
    \end{align*} and due to \begin{align*}
    \frac{b(c)}{b(1/p(\ell_c))} = \frac{b(c)}{b(c/(p(\ell_c)c))} \sim (p(\ell_c) \cdot c)^{1/\beta} \rightarrow \lambda^{1/\beta} \text{ as } c \rightarrow \infty.
    \end{align*} the second factor converges to \(\lambda^{1/\beta}\),
it follows that \(Z \sim ML(\beta,1)\). Since \(c \rightarrow \infty\)
is equivalent to \(\ell \uparrow x_R\) the assertion follows with
\(\ell:=\ell_c\) and \(p_\ell:=p(\ell_c)\). \qed

@@

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka, 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-Moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computationally intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes.

@@

\caption{\label{fig:MSE} MSE for the estimation of tail (left column) and scale (right column) parameters via log-Moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one, based on 5000 simulation runs.}\label{fig:MSE}
\end{figure}

Since the Mittag-Leffler distribution is heavy-tailed, many researchers
would intuitively give the highest importance to the tail behaviour of
the distribution. Of course, one can also use established tail exponent
estimators for the estimation of the parameter \(\beta\) such as the
Hill estimator (Hill, 1975) based on the \(r+1\) upper order statistics
\begin{align}\label{eq:Hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align} where \(X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}\)
denote the order statistics in decreasing order of a sequence
\(X_1,\ldots,X_n\). The Hill estimator only performs reliably well for
distributions close to Pareto (see Resnick, 1997). It heavily depends on
the choice of \(r\) in \eqref{eq:Hill}, and hence on the number of order
statistics entering \eqref{eq:Hill}. Figure \ref{fig:Hillplots} shows
Hill plots for simulated Mittag-Leffler data, with varying sample sizes
and tails. To deduce the correct tail parameter estimates from one of
those Hillplots is virtually impossible. To estimate the tail parameter
of a Mittag-Leffler distribution with the Hill estimator, the sample
size has to be much larger and one has to choose a small tuning
parameter \(r\). In case that only events above a high threshold are
recorded, \(n=10000\) events are extremely unrealistic. There is a vast
literature on bias-reduced versions of the Hill estimator and on the
optimal choice of \(r\), mainly based on imposed second-order regular
variation conditions (see e.g.~Beirlant et al., 2006, p. Ch.4.4 and 4.7
and references in it). Also, there are hundreds other tail-index
estimators not considered in this article, since this would go beyond
the scope of our article. For example the moment estimator (Dekkers et
al., 1989), the Pickands estimator (Pickands III and others, 1975), the
QQ-estimator (Kratz and Resnick, 1996), to name a few. A nice overview
can be found in (Fedotenkov, 2018). However, we know from the Theorem in
Section \ref{sec:scaling} that the inter-exceedance times are
asymptotically Mittag-Leffler distributed, and hence it is unsurprising
that general tail (and scale) estimation methods that do not take into
account this structure are statistically less efficient (Kozubowski,
2001). Finally, the Hill estimator completely fails in the important
special case where the inter-arrival times are exponentially distributed
and not heavy-tailed. For these reasons, we propose to use the
log-Moment estimator or the MLE for Mittag-Leffler distributions.

@@

\caption{\label{fig:Hillplots} Hillplots for m=100 simulated Mittag-Leffler datasets with true tail 0.8 and sample size 200 (left) and 10000 (right), with number of upper order statistics r on which the Hill estimator is based on the x-axis. The grey thin lines are the Hill plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter. }\label{fig:Hillplots}
\end{figure}

@@

  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    extract the set \(\mathcal T_k\) of exceedance times between the
    magnitudes exceeding the threshold \(X_{(k)}\)
  \item
    fit a Mittag-Leffler distribution to \(\mathcal T_k\), resulting in
    the parameter estimates \(\hat\beta_k\) and \(\hat \sigma_k\).
  \end{enumerate}
\item
  Plot \(k\) vs.~\(\hat \beta_k\). To the right (\(k \uparrow K\), low
  threshold), the asymptotics are off and bias is high. To the left
  (\(k \downarrow K_0\), high threshold), data is scarce and variance is
  high. In the middle, look for a region of \emph{stability} for a
  parameter estimate \(\hat \beta\). Choose as \(\hat \beta\) a
  representative value from this region.
\item
  Plot \(k\) vs.~\(k^{1/\hat \beta} \hat \sigma_k\). Again, choose a
  region of \emph{stability} and, and choose a representative
  \(\hat \sigma_0\) from that region.
\end{enumerate}

@@

In order to have exact analytical values available for \(\beta\) and
\(\sigma_0\), a distribution for \(W_k\) needs to be chosen for which
\(b(n)\) from \eqref{eq:stability} is known.

@@

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/hazard-1} \caption{\label{fig:hazard} Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\beta = 0.8$, $\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\beta$.}\label{fig:hazard}
\end{figure}

>>>>

Often a magnitude, or mark can be assigned to each event in the renewal
process, such as for earthquakes, solar flares or neuron voltages. The
Peaks-Over-Threshold model (POT, see e.g.~Coles, 2001) applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution
to the threshold exceedances. A commonly made assumption in POT models
is that times between events are either fixed or light-tailed, and this
entails that the threshold crossing times form a Poisson process (``On
the Exceedance Point Process for a Stationary Sequence,'' 1988). Then as
one increases the threshold \(\ell\) and thus decreases the threshold
crossing probability \(p_{\ell}\), the Poisson process is rarefied,
i.e.~its intensity decreases \emph{linearly} with \(p_{\ell}\) (see
e.g.~Beirlant et al., 2006).

As will be shown below, in the heavy-tailed waiting time scenario
threshold crossing times form a \emph{fractional Poisson process}
(Laskin, 2003; Meerschaert et al., 2011), which is a renewal process
with Mittag-Leffler distributed waiting times. The family of
Mittag-Leffler distributions nests the exponential distribution (Haubold
et al., 2011), and hence the fractional Poisson process generalizes the
standard Poisson process. Again as the threshold size \(\ell\) increases
and the threshold crossing probability \(p_{\ell}\) decreases, the
fractional Poisson process is rarefied: The scale parameter of the
Mittag-Leffler inter-arrival times of threshold crossing times
increases, but \emph{superlinearly}; see the Theorem below.

@@

In the following we denote with \(x_L \in [-\infty, +\infty)\) and
\(x_R \in (-\infty, +\infty]\) the left and right endpoint of the
distribution of \(J\). We assume that the \(k\)-th magnitude \(J_k\)
occurs at time \(T_k = W_1 + \ldots + W_k\). Based on a MRP, we define
the Continuous Time Random Exceedance model (CTRE) as follows:

@@

\begin{description}
\item[\textbf{Theorem:}]
For the magnitudes \(J_k\) assumption \eqref{assumptionJs} holds.
Furthermore, let the waiting times \(W_k\) be in the domain of
attraction of a positively skewed sum-stable law with stability
parameter \(0 < \beta < 1\); more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align} for a function \(b(n)\) which is regularly varying at
\(\infty\) with parameter \(1/\beta\), and where
\(\mathbf E[\exp(-sD)] = \exp(-s^\beta)\). Write
\(p_{\ell} := \mathbf P(J > \ell)\). Then the weak convergence \[
\frac{T(\ell)} {b(1/p_{\ell})} \to Z_\beta \quad \text{ as } \quad \ell \uparrow x_R
\] holds, where the Mittag-Leffler random variable \(Z_\beta\) is
defined on the positive real numbers via \[
\mathbf E[\exp(-sZ_\beta)] = \frac{1}{1+s^\beta}.
\]
\end{description}

\emph{Proof of Theorem:} Due to assumption \eqref{assumptionJs}, it
follows \begin{align*}
    P\left(\frac{M(c)-d(c)}{a(c)} \leq x\right) \rightarrow F(x) \text{ as } c \rightarrow \infty,
    \end{align*} where \(x\) is any value from the support of \(F\).
Then it follows, \begin{align*}
      F(xa(c)+d(c))^{\lfloor c \rfloor} &\rightarrow F(x) \text{ as } c \rightarrow \infty,\\
      \Rightarrow  \lfloor c \rfloor \log F(\ell_c)  &\rightarrow \log F(x)\text{ as } c \rightarrow \infty,
    \end{align*} with \(\ell_c:=xa(c)+d(c)\). Furthermore, since
\(\log(1-x) \sim -x\) for small \(x\) it follows \begin{align}
      c \cdot p(\ell_c) \rightarrow -\log F_A(x)  \text{ as } c \rightarrow \infty,\label{eq:lplconv}
    \end{align} with \(p(\ell_c):= 1-F(\ell_c)\). Since
\(\tau(\ell_c) \sim Geo(p(\ell_c))\) and \eqref{eq:lplconv} it follows
that \(\tau(\ell_c)/c\) converges to an Exponential distributed random
variable, \begin{align*}
    \frac{\tau(\ell_c)}{c} \Rightarrow E  \text{ as } c \rightarrow \infty
    \end{align*} with inverse mean \(\lambda:=-\log F_A(x)\). Due to
\begin{align*}
    S(c):= \sum_{i=1}^{\lfloor c \rfloor} \frac{W_i}{b(c)} \Rightarrow D \ \text{ as } c \rightarrow \infty,
    \end{align*} it follows with Gnedenkos transfer theorem (see
Gnedenko (1983)), that \begin{align}
    \sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(\ell_c)}  \Rightarrow Z\ \text{ as }  c \rightarrow \infty, \label{TheoGutArg}
    \end{align} where the distribution of \(Z\) has the characteristic
function \begin{align}
    \Psi(s) &= \int_0^{\infty} (\Phi_D(s))^{y} F_E(dy) \nonumber\\
    &= \int_0^{\infty} \frac{1}{1-\log(\Phi_D(s))/\lambda} \nonumber\\
    &= \frac{1}{1+s^{\beta}\lambda^{-1}}
    \end{align} where \(\Phi_D(s) = \exp(-s^\beta)\) is the Laplace
transform of \(D\) and \(F_E(dy)\) is the distribution function of
\(E\). Hence, \(Z\) is Mittag-Leffler distributed with parameter
\(\lambda^{-1/\beta}\). Rewriting \begin{align*}
    \sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(1/p(\ell_c))} = \left(\sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(c) }\right) \frac{b(c)}{b(1/p(\ell_c))}  
    \end{align*} and \begin{align*}
    \frac{b(c)}{b(1/p(\ell_c))} = \frac{b(c)}{b(c/(p(\ell_c)c))} \sim (p(\ell_c) \cdot c)^{1/\beta} \rightarrow \lambda^{1/\beta} \text{ as } c \rightarrow \infty,
    \end{align*} the second factor converges to \(\lambda^{1/\beta}\),
and it follows that \(Z \sim ML(\beta,1)\). Since
\(c \rightarrow \infty\) is equivalent to \(\ell \uparrow x_R\), the
assertion follows with \(\ell:=\ell_c\) and \(p_\ell:=p(\ell_c)\). \qed

@@

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka, 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-Moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computationally intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes. For smaller tails both estimators show an increasing
rmse for the scale estimation due to an increasing variance. This
results from the fact that in case of very small tails, single very
large values can occur.

@@

\caption{\label{fig:MSE} Rmse for the estimation of tail (left column) and scale (right column) parameters via log-Moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one, based on 5000 simulation runs.}\label{fig:MSE}
\end{figure}

Since the Mittag-Leffler distribution is heavy-tailed, many researchers
would intuitively give the highest importance to the tail behaviour of
the distribution. Of course, one can also use established tail exponent
estimators for the estimation of the parameter \(\beta\) such as the
Hill estimator (Hill, 1975) based on the \(r+1\) upper order statistics
\begin{align}\label{eq:Hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align} where \(X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}\)
denote the order statistics in decreasing order of a sequence
\(X_1,\ldots,X_n\). The Hill estimator only performs reliably well for
distributions close to Pareto (see Resnick, 1997). It heavily depends on
the choice of \(r\) in \eqref{eq:Hill}, and hence on the number of order
statistics entering \eqref{eq:Hill}. Figure \ref{fig:Hillplots} shows
100 Hill plots (grey thin lines) for simulated Mittag-Leffler data with
tail parameter \(\beta=0.8\) and scale \(\sigma=1\), for a sample size
of \(n=200\) (left) and \(n=1000\) (right). The dark lines are the means
and the red dotted lines show the true tail parameter. To deduce the
correct tail parameter estimates from one of those Hillplots is
virtually impossible. To estimate the tail parameter of a Mittag-Leffler
distribution with the Hill estimator, the sample size has to be much
larger and one has to choose a small tuning parameter \(r\). In case
that only events above a high threshold are recorded, \(n=10000\) events
are extremely unrealistic. There is a vast literature on bias-reduced
versions of the Hill estimator and on the optimal choice of \(r\),
mainly based on imposed second-order regular variation conditions (see
e.g.~Beirlant et al., 2006, p. Ch.4.4 and 4.7 and references in it).
Also, there are hundreds other tail-index estimators not considered in
this article, since this would go beyond the scope of our article. For
example the moment estimator (Dekkers et al., 1989), the Pickands
estimator (Pickands III and others, 1975), the QQ-estimator (Kratz and
Resnick, 1996), to name a few. A nice overview can be found in
(Fedotenkov, 2018). However, we know from the Theorem in Section
\ref{sec:scaling} that the inter-exceedance times are asymptotically
Mittag-Leffler distributed, and hence it is unsurprising that general
tail (and scale) estimation methods that do not take into account this
structure are statistically less efficient (Kozubowski, 2001). Finally,
the Hill estimator completely fails in the important special case where
the inter-arrival times are exponentially distributed and not
heavy-tailed. For these reasons, we propose to use the log-Moment
estimator or the MLE for Mittag-Leffler distributions.

@@

\caption{\label{fig:Hillplots} Hillplots for m=100 simulated Mittag-Leffler datasets with true tail 0.8 and sample size 200 (first panel) and 10000 (second panel), with number of upper order statistics r on which the Hill estimator is based on the x-axis. The grey thin lines are the Hill plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter. }\label{fig:Hillplots}
\end{figure}

@@

  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    extract the set \(\mathcal T_k\) of exceedance times between the
    magnitudes exceeding the threshold \(X_{(k)}\)
  \item
    fit a Mittag-Leffler distribution to \(\mathcal T_k\), resulting in
    the parameter estimates \(\hat\beta_k\) and \(\hat \sigma_k\).
  \end{enumerate}
\item
  Plot \(k\) vs.~\(\hat \beta_k\). To the right (\(k \uparrow K\), low
  threshold), the asymptotics are off and bias is high. To the left
  (\(k \downarrow K_0\), high threshold), data is scarce and variance is
  high. In the middle, look for a region of \emph{stability} for a
  parameter estimate \(\hat \beta\). Choose as \(\hat \beta\) a
  representative value from this region.
\item
  Plot \(k\) vs.~\(k^{1/\hat \beta} \hat \sigma_k\). Again, choose a
  region of \emph{stability} and a representative \(\hat \sigma_0\) from
  that region.
\end{enumerate}

@@

The magnitudes are in all scenarios standard Gumbel distributed. In
order to have exact analytical values available for \(\beta\) and
\(\sigma_0\), a distribution for \(W_k\) needs to be chosen for which
\(b(n)\) from \eqref{eq:stability} is known.

@@

\begin{figure}
\includegraphics[width=\textwidth]{article_springer_files/figure-latex/hazard-1} \caption{\label{fig:hazard} Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\beta = 0.9$, $\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\beta$.}\label{fig:hazard}
\end{figure}
